---
output: github_document
---

# ExpoMet

<!-- README.md is generated from README.Rmd. Please edit README.Rmd, not README.md. -->

`ExpoMet` is an R package designed to make common exposomic and high-dimensional omic analyses accessible to epidemiologists.

The package provides tools for:

- exploratory analysis and visualization of exposomic data;
- handling values below the limit of detection;
- transforming high-dimensional features;
- exposure-wide association analyses;
- Cox proportional hazards regression;
- mixed-effects regression;
- preparation of results for downstream pathway analysis.

## Installation

`ExpoMet` is currently distributed through GitHub and is not available on CRAN.

### Install directly from GitHub

```{r install-github, eval=FALSE}
install.packages("remotes")

remotes::install_github("haotianwu17/ExpoMet")
```

Then load the package:

```{r load-package, eval=FALSE}
library(ExpoMet)
```

### Install from a downloaded source package

A `.tar.gz` source package can also be downloaded from the GitHub Releases page and installed locally.

```{r install-source, eval=FALSE}
install.packages(
  "ExpoMet_0.1.0.tar.gz",
  repos = NULL,
  type = "source"
)
```

For example, on Windows:

```{r install-source-windows, eval=FALSE}
install.packages(
  "C:/Users/YourName/Downloads/ExpoMet_0.1.0.tar.gz",
  repos = NULL,
  type = "source"
)
```

Replace the version number and file path as appropriate.

## Main functions

The current ExpoMet workflow includes the following functions:

| Function | Purpose |
|---|---|
| `preview()` | Preview and summarize exposomic data before analysis |
| `detection()` | Evaluate detection patterns of exposomic features |
| `replace_below_lod()` | Replace measurements below the limit of detection |
| `transform_features()` | Apply transformations to exposomic features |
| `exp_eda()` | Perform exploratory data analysis of exposomic features |
| `run_regression()` | Run exposure-wide regression analyses |
| `run_cox()` | Run exposure-wide Cox proportional hazards models |
| `run_mixed()` | Run exposure-wide mixed-effects models |
| `format_mummichog_input()` | Format association results for downstream mummichog analysis |

Detailed documentation for each function is available in R using:

```{r help-functions, eval=FALSE}
?preview
?detection
?replace_below_lod
?transform_features
?exp_eda
?run_regression
?run_cox
?run_mixed
?format_mummichog_input
```

## Example data

ExpoMet includes example datasets that can be used to learn the workflow.

```{r example-data, eval=FALSE}
library(ExpoMet)

data(exposome_example)
data(pheno_example)

dim(exposome_example)
head(pheno_example)
```

`exposome_example` contains participant IDs and simulated exposomic features.

`pheno_example` contains participant-level phenotype and covariate information that can be linked to the exposomic data using `exp_id`.

## Quick example: exposure-wide association analysis

For a binary outcome, first ensure that the outcome is coded appropriately for binomial regression.

```{r regression-example, eval=FALSE}
library(ExpoMet)

data(exposome_example)
data(pheno_example)

pheno_example$Disease_binary <- ifelse(
  pheno_example$Disease == "Yes",
  1,
  0
)

fit <- run_regression(
  Disease_binary ~ omic_features + age + sex,
  pheno = pheno_example,
  omic_features = exposome_example,
  id_col = "exp_id",
  family = "binomial"
)

head(fit$results)
```

In the model formula, `omic_features` acts as a placeholder. ExpoMet fits the specified model separately for each exposomic feature while retaining the other variables in the formula as covariates.

Results can be accessed using:

```{r regression-results, eval=FALSE}
fit$results
```

## Typical ExpoMet workflow

A general analysis may proceed as follows:

```{r workflow, eval=FALSE}
library(ExpoMet)

# 1. Load exposomic and phenotype data
data(exposome_example)
data(pheno_example)

# 2. Inspect the exposomic dataset
preview(exposome_example)

# 3. Evaluate detection
# See ?detection for available arguments

# 4. Handle measurements below the limit of detection
# See ?replace_below_lod

# 5. Transform features if needed
# See ?transform_features

# 6. Explore distributions and data structure
# See ?exp_eda

# 7. Run association models
# See ?run_regression, ?run_cox, and ?run_mixed

# 8. Prepare results for pathway analysis if needed
# See ?format_mummichog_input
```

The exact preprocessing steps required will depend on the study design, exposure platform, and scientific question.

## Regression models

### Generalized linear models

Use `run_regression()` for exposure-wide generalized linear regression, including continuous, binary, and count outcomes as supported by the function.

```{r glm-example, eval=FALSE}
fit <- run_regression(
  Disease_binary ~ omic_features + age + sex,
  pheno = pheno_example,
  omic_features = exposome_example,
  id_col = "exp_id",
  family = "binomial"
)
```

### Cox proportional hazards models

Use:

```{r cox-help, eval=FALSE}
?run_cox
```

for time-to-event analyses.

The `survival` package is an optional dependency and is required when using Cox regression functionality.

If needed:

```{r install-survival, eval=FALSE}
install.packages("survival")
```

### Mixed-effects models

Use:

```{r mixed-help, eval=FALSE}
?run_mixed
```

for analyses requiring mixed-effects models.

The `lme4` package is an optional dependency and is required for the corresponding mixed-model functionality.

If needed:

```{r install-lme4, eval=FALSE}
install.packages("lme4")
```

## Reproducibility

For reproducible analyses, we recommend recording the versions of R and ExpoMet used in an analysis:

```{r session-info, eval=FALSE}
packageVersion("ExpoMet")
sessionInfo()
```

## Development

ExpoMet is under active development.

Issues, bug reports, and suggestions can be submitted through the GitHub repository:

<https://github.com/haotianwu17/ExpoMet>

## Authors

ExpoMet is developed collaboratively by the ExpoMet team.
